Probabilistic Model of Bilateral Lymphatic Spread in Head and Neck Cancer

### 2024-05-28 _tex/manuscript.tex
Built site for gh-pages

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  letterpaper,
  DIV=11,
  numbers=noendperiod]{scrartcl}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{multirow}
\usepackage{centernot}
\KOMAoption{captions}{tableheading}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

@@

\title{Probabilistic Model of Bilateral Lymphatic Spread in Head and
Neck Cancer}
\author{Roman Ludwig \and Yoel Perez Haas \and Jan Unkelbach}
\date{}

@@

\caption{}\label{T_cf4d9}\tabularnewline
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Centre Léon Bérard & 325 & 60 & 100\% & 19\% & 69\% & 18\% \\
Inselspital Bern & 74 & 61 & 100\% & 18\% & 66\% & 14\% \\
University Hospital Zurich & 287 & 66 & 26\% & 18\% & 52\% & 31\% \\
Vall d\textquotesingle Hebron Barcelona Hospital & 147 & 58 & 5\% & 21\%
& 34\% & 34\% \\

@@

\subsection{Requirements for a Bilateral
Model}\label{requirements-for-a-bilateral-model}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  More advanced T-category should lead to higher risk for nodal disease.
  One approach to achieve this via the expected time of diagnosis has
  already been developed in the form of a hidden Markov model (Ludwig et
  al. 2021).
\item
  The degree of ipsilateral involvement should give the model
  information on the time that may have passed between onset and
  diagnosis of the disease. This should come in addition to what can be
  inferred about this time from T-category alone.
\item
  A tumor that extends over the mid-sagittal line should yield
  contralateral metastases with much higher probability.
\end{enumerate}

@@

The entire data, including additional patients with tumors in other
primary locations than the oropharynx, is publicly available: It may be
\href{https://lyprox.org/patients/dataset}{downloaded from LyProX} where
it can be interactively explored too,
\href{https://github.com/rmnldwg/lydata}{from GitHub},
\href{https://zenodo.org/search?q=lydata}{from zenodo}, or via the
\emph{Data-in-Brief} publications Ludwig et al. (2022) and Ludwig et al.
(2023). Although these publications do not include the most recent
dataset addition from Vall d'Hebron Barcelona Hospital.

@@

\includegraphics[width=0.6\textwidth,height=\textheight]{static/small-graph.png}

@@

Our first model to predict the lymphatic progression of HNSCC was
introduced using Bayesian networks (Pouymayou et al. 2019). We
subsequently extended this work to a hidden Markov model (HMM) (Ludwig
et al. 2020) to allow an intuitive inclusion of T-category into the
predictions. We will briefly summarize this HMM's formalism before
building on it to include the contralateral spread in
section~\ref{sec-ext-to-contra}.

@@

We compared three different aprroaches to modelling the contralateral
involvement in OPSCC:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  A naive model of the contralateral side only. Neither the patterns of
  ipsilateral involvement, nor the information on whether the patients'
  tumors crossed the mid-sagittal line enter this model.
\item
  The bilateral model as introduced in the sections up to and including
  section~\ref{sec-params-symmetry}. The involvement patterns of both
  sides enter this model, but not whether the tumors extended the
  midline.
\item
  An additional bilateral model that tracks the midline extension of the
  patients' tumors as a random variable. This includes the formalism
  introduced up to and including section~\ref{sec-midline}. In contrast
  to the other two models, it does consider the reported midline
  extension in the data to attempt a more accurate prediction of the
  risk for contralateral occult disease.
\end{enumerate}

@@

\begin{itemize}
\tightlist
\item
  all models initialized with the same settings

  \begin{itemize}
  \tightlist
  \item
    same graph
  \item
    same data
  \item
    same distributions over diagnose time
  \end{itemize}
\item
  parameter samples were drawn for all three models:

  \begin{itemize}
  \tightlist
  \item
    the sampling used 12 walkers per parameter space dimension.
  \item
    Every 50 steps, convergence was checked
  \item
    convergence was reached when two criteria were met:

    \begin{enumerate}
    \def\labelenumi{\arabic{enumi}.}
    \tightlist
    \item
      the relative difference between subsequent autocorrelation
      estimates of the chain does not change more than 5.0e-2.
    \item
      the autocorrelation estimate of the chain drops below the length
      of the chain divided by 50.
    \end{enumerate}
  \item
    after convergence, 10 samples (from the parallel walkers) were drawn
    that were spaced 10 steps apart, resulting in
    \texttt{num\ params\ x\ 12\ x\ 10} parameter samples.
  \end{itemize}
\end{itemize}

For every model, we compare the predicted prevalence of a certain
scenario to how often this scenario is seen in the data. The considered
scenarios are choosen to test if the models can take into account
T-category, the correlation of ipsi- and contralateral involvement, and
the primary tumor's lateralization (whether or not it extends over the
mid-sagittal line).

\begin{itemize}
\tightlist
\item[$\square$]
  list and describe the scenarios
\end{itemize}

@@

@@

\caption{\label{fig-model-burnin-history}Monitoring quantities during
the burn-in phase of the parameter sampling.}

@@

figure~\ref{fig-model-burnin-history} is only a temporary placeholder
for now.

@@

@@

\subsubsection{Midline Extension}\label{midline-extension}

@@

\subsection{Parameter Estimates}\label{parameter-estimates}

\caption{}\label{T_17196}\tabularnewline
\toprule\noalign{}
Parameter & Mean & Std. Dev. \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Parameter & Mean & Std. Dev. \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Mid. ext. probability & 8.13\% & ± 0.60\% \\
ipsi: T ➜ II & 35.45\% & ± 1.75\% \\
ipsi: T ➜ III & 5.59\% & ± 0.82\% \\
ipsi: T ➜ IV & 0.91\% & ± 0.21\% \\
contra: T ➜ II & 2.64\% & ± 0.35\% \\
contra: T ➜ III & 0.16\% & ± 0.09\% \\
contra: T ➜ IV & 0.21\% & ± 0.10\% \\
Mixing ⍺ & 21.01\% & ± 3.24\% \\
II ➜ III & 13.46\% & ± 1.95\% \\
III ➜ IV & 15.74\% & ± 2.24\% \\
late T-cat. binom. prob. & 45.35\% & ± 2.57\% \\

The predicted prevalence of midline extension for early T-category
tumors is . For a late T-category tumor, it is .

In the data, the prevalence of midline extension for early T-category
tumors is . For late T-category tumors it is .

\begin{itemize}
\tightlist
\item
  Discrepancy between data and model is due to late T-stage's time prior
\item
  still tied to other parameters
\item
  cannot fully represent spread of midline extension and also represent
  spread params
\end{itemize}

@@

\caption{}\label{T_e13a6}\tabularnewline
\toprule\noalign{}
T-cat. & ipsi & Mid. ext. & \multicolumn{2}{l}{%
I} & \multicolumn{2}{l}{%
II} & \multicolumn{2}{l}{%
III} & \multicolumn{2}{l}{%
IV} & total \\
& & & n & \% & n & \% & n & \% & n & \% & n \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
T-cat. & ipsi & Mid. ext. & \multicolumn{2}{l}{%
I} & \multicolumn{2}{l}{%
II} & \multicolumn{2}{l}{%
III} & \multicolumn{2}{l}{%
IV} & total \\
& & & n & \% & n & \% & n & \% & n & \% & n \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
early & 0 & False & 0 & 0.00 & 1 & 1.16 & 0 & 0.00 & 0 & 0.00 & 86 \\
early & 0 & True & 0 & 0.00 & 1 & 10.00 & 1 & 10.00 & 0 & 0.00 & 10 \\
early & 1 & False & 1 & 0.53 & 11 & 5.82 & 2 & 1.06 & 1 & 0.53 & 189 \\
early & 1 & True & 1 & 11.11 & 2 & 22.22 & 0 & 0.00 & 0 & 0.00 & 9 \\
early & ≥ 2 & False & 1 & 0.96 & 15 & 14.42 & 3 & 2.88 & 3 & 2.88 &
104 \\
early & ≥ 2 & True & 0 & 0.00 & 3 & 30.00 & 4 & 40.00 & 1 & 10.00 &
10 \\
advanced & 0 & False & 0 & 0.00 & 2 & 5.88 & 0 & 0.00 & 0 & 0.00 & 34 \\
advanced & 0 & True & 0 & 0.00 & 3 & 12.50 & 0 & 0.00 & 0 & 0.00 & 24 \\
advanced & 1 & False & 0 & 0.00 & 3 & 4.55 & 0 & 0.00 & 0 & 0.00 & 66 \\
advanced & 1 & True & 1 & 1.64 & 18 & 29.51 & 5 & 8.20 & 1 & 1.64 &
61 \\
advanced & ≥ 2 & False & 4 & 5.26 & 16 & 21.05 & 7 & 9.21 & 4 & 5.26 &
76 \\
advanced & ≥ 2 & True & 3 & 3.80 & 46 & 58.23 & 18 & 22.78 & 8 & 10.13 &
79 \\

@@

\phantomsection\label{refs}
\begin{CSLReferences}{1}{0}
\bibitem[\citeproctext]{ref-ludwig_dataset_2022}
Ludwig, Roman, Jean-Marc Hoffmann, Bertrand Pouymayou, Grégoire Morand,
Martina Broglie Däppen, Matthias Guckenberger, Vincent Grégoire,
Panagiotis Balermpas, and Jan Unkelbach. 2022. {``A Dataset on
Patient-Individual Lymph Node Involvement in Oropharyngeal Squamous Cell
Carcinoma.''} \emph{Data in Brief} 43 (August): 108345.
\url{https://doi.org/10.1016/j.dib.2022.108345}.

\bibitem[\citeproctext]{ref-ludwig_hidden_2020}
Ludwig, Roman, Bertrand Pouymayou, Panagiotis Balermpas, and Jan
Unkelbach. 2020. {``A {Hidden Markov} Model for Lymphatic Tumour
Progression.''} In \emph{{SASRO}}.

\bibitem[\citeproctext]{ref-ludwig_hidden_2021}
---------. 2021. {``A Hidden {Markov} Model for Lymphatic Tumor
Progression in the Head and Neck.''} \emph{Sci Rep} 11 (1): 12261.
\url{https://doi.org/10.1038/s41598-021-91544-1}.

\bibitem[\citeproctext]{ref-ludwig_multicentric_2023}
Ludwig, Roman, Adrian Schubert, Dorothea Barbatei, Laurence Bauwens,
Sandrine Werlen, Olgun Elicin, Matthias Dettmer, et al. 2023. {``A
Multi-Centric Dataset on Patient-Individual Pathological Lymph Node
Involvement in Head and Neck Squamous Cell Carcinoma.''} \emph{Data in
Brief}, December, 110020.
\url{https://doi.org/10.1016/j.dib.2023.110020}.

\bibitem[\citeproctext]{ref-pouymayou_bayesian_2019}
Pouymayou, Bertrand, Panagiotis Balermpas, Oliver Riesterer, Matthias
Guckenberger, and Jan Unkelbach. 2019. {``A {Bayesian} Network Model of
Lymphatic Tumor Progression for Personalized Elective {CTV} Definition
in Head and Neck Cancers.''} \emph{Physics in Medicine \& Biology} 64
(16): 165003. \url{https://doi.org/10.1088/1361-6560/ab2a18}.

\end{CSLReferences}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  sn-mathphys-num,
]{sn-jnl}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

%%%% Standard Packages

\usepackage{graphicx}%
\usepackage{multirow}%
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{amsthm}%
\usepackage{mathrsfs}%
\usepackage[title]{appendix}%
\usepackage{xcolor}%
\usepackage{textcomp}%
\usepackage{manyfoot}%
\usepackage{booktabs}%
\usepackage{algorithm}%
\usepackage{algorithmicx}%
\usepackage{algpseudocode}%
\usepackage{listings}%

%%%%

\raggedbottom
\usepackage{multirow}
\usepackage{centernot}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{plainnat}
\usepackage{bookmark}

@@

\title[Probabilistic Model of Bilateral Lymphatic Spread in Head and
Neck Cancer]{Probabilistic Model of Bilateral Lymphatic Spread in Head
and Neck Cancer}

% author setup
\author*[1,2]{\fnm{Roman} \sur{Ludwig}}\email{roman.ludwig@usz.ch}\author[1,2]{\fnm{Yoel Perez} \sur{Haas}}\email{yoel.perezhaas@usz.ch}\author[1,2]{\fnm{Jan} \sur{Unkelbach}}\email{jan.unkelbach@usz.ch}
% affil setup
\affil[1]{\orgdiv{Department of Physics}, \orgname{University of
Zurich}}
\affil[2]{\orgdiv{Radiation Oncology}, \orgname{University Hospital
Zurich}}

% abstract

% keywords

@@

\caption{}\label{T_900f1}\tabularnewline
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Institution & Total & Age (median) & Neck Dissection & N0 & Early T-Cat.
& Mid. Ext. \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Centre Léon Bérard & 325 & 60 & 100\% & 19\% & 69\% & 18\% \\
Inselspital Bern & 74 & 61 & 100\% & 18\% & 66\% & 14\% \\
University Hospital Zurich & 287 & 66 & 26\% & 18\% & 52\% & 31\% \\
Vall d\textquotesingle Hebron Barcelona Hospital & 147 & 58 & 5\% & 21\%
& 34\% & 34\% \\

@@

\subsection{Requirements for a Bilateral Model}\label{sec-requirements}

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  More advanced T-category should lead to higher risk for nodal disease.
  One approach to achieve this via the expected time of diagnosis has
  already been developed in the form of a hidden Markov model
  \citep{ludwig_hidden_2021}.
\item
  The degree of ipsilateral involvement should give the model
  information on the time that may have passed between onset and
  diagnosis of the disease. This should come in addition to what can be
  inferred about this time from T-category alone.
\item
  A tumor that extends over the mid-sagittal line should yield
  contralateral metastases with much higher probability.
\end{enumerate}

@@

The entire data, including additional patients with tumors in other
primary locations than the oropharynx, is publicly available: It may be
\href{https://lyprox.org/patients/dataset}{downloaded from LyProX} where
it can be interactively explored too,
\href{https://github.com/rmnldwg/lydata}{from GitHub},
\href{https://zenodo.org/search?q=lydata}{from zenodo}, or via the
\emph{Data-in-Brief} publications \citet{ludwig_dataset_2022} and
\citet{ludwig_multicentric_2023}. Although these publications do not
include the most recent dataset addition from Vall d'Hebron Barcelona
Hospital.

@@

\includegraphics[width=0.5\textwidth,height=\textheight]{static/small-graph.png}

@@

Our first model to predict the lymphatic progression of HNSCC was
introduced using Bayesian networks \citep{pouymayou_bayesian_2019}. We
subsequently extended this work to a hidden Markov model (HMM)
\citep{ludwig_hidden_2021} to allow an intuitive inclusion of T-category
into the predictions. We will briefly summarize this HMM's formalism
before building on it to include the contralateral spread in
section~\ref{sec-ext-to-contra}.

@@

In this section, we detail how the experiments were performed. Every
figure, table, and result is fully reproducible via the GitHub
repository
\href{https://github.com/rmnldwg/bilateral-paper}{\texttt{rmnldwg/bilateral-paper}}.
It also contains the raw manuscript and instructions on how to recreate
all figures, tables, and the final document.

\subsection{Involvement Data
Consensus}\label{involvement-data-consensus}

It is possible to provide our model with multiple different diagnostic
modalities, each being characterized by different pairs of sensitivity
and specificity. However, we instead chose to combine them into a single
``consensus'' diagnosis before parameter inference. We opted for this
because the literature values of sensitivity and specificity
\citep{debondt_detection_2007} of imaging modalities like MRI and CT do
not plausibly match some of our observations: In the USZ cohort, 78\% of
OPSCC patients where diagnosed with ipsilateral LNL II involvement via
diagnostic imaging. This is virtually impossible with sensitivities
around 80\% and specificities lower than 100\%.

Our pre-training consensus was formed by considering all reported
diagnostic information for a particular patient and LNL. When conflicts
arose, we computed the \emph{most likely} true state of involvement
using the literature sensitivity and specificity values
\citep{debondt_detection_2007}.

@@

For parameter inference, we used the Python package
\href{https://emcee.readthedocs.io/en/stable/}{\texttt{emcee}}
\citep{foreman-mackey_emcee_2013}. It implements efficient MCMC sampling
algorithms that employ multiple parallel samplers for affine invariance
and better performance on multi-core CPUs. The
\href{https://emcee.readthedocs.io/en/stable/}{\texttt{emcee}} library
was provided with the likelihood implemented by our
\href{https://lymph-model.readthedocs.io/en/stable/}{\texttt{lymph-model}}
Python package.

For each dimension in the parameter space of the model, we initialized
12 of these parallel samplers, called ``walkers'', with random values in
the unit cube. Every time all of these walkers advanced 50 steps, the
autocorrelation time of the chains was estimated. For short chains, this
estimate is not trustworthy, but stabilizes for longer chains. We
therefore considered a sampling to be converged when two criteria where
met:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  The change in the autocorrelation time was less then 5.0e-2.
\item
  The estimate of the autocorrelation dropped below \(n\) / 50 where
  \(n\) is the length of the chain up to that point.
\end{enumerate}

All samples up to this convergence - called the \emph{burn-in phase} -
were discarded. We only kept another 10 samples after that, which were
spaced 10 steps apart.

@@

First, in figure~\ref{fig-model-burnin-history}, we verify the sampling
converged successfully by inspecting two monitoring quantities: The
autocorrelation time of the MCMC chain and the acceptance fractions of
the parallel walkers.

@@

\caption{\label{fig-model-burnin-history}Monitoring quantities during
the burn-in phase of the parameter sampling. Left: The autocorrelation
time of the sampling chain estimated at different sampling steps. We
consider the chain converged when the estimate of the autocorrelation
time is stable and drops below the trust threshold of \(n/50\) where
\(n\) is the number of steps. Right: Fraction of accepted MCMC proposals
averaged over all parallel walkers. Values around 30\% indicate good
mixing of the walkers.}

@@

In table~\ref{tbl-midline-params}, we tabulate the mean and standard
deviation of the sampled parameters for the full midline model.

\caption{}\label{T_4992d}\tabularnewline
\toprule\noalign{}
Parameter & Mean & Std. Dev. \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
Parameter & Mean & Std. Dev. \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Mid. ext. probability & 8.13\% & ± 0.60\% \\
ipsi: T ➜ II & 35.45\% & ± 1.75\% \\
ipsi: T ➜ III & 5.59\% & ± 0.82\% \\
ipsi: T ➜ IV & 0.91\% & ± 0.21\% \\
contra: T ➜ II & 2.64\% & ± 0.35\% \\
contra: T ➜ III & 0.16\% & ± 0.09\% \\
contra: T ➜ IV & 0.21\% & ± 0.10\% \\
Mixing ⍺ & 21.01\% & ± 3.24\% \\
II ➜ III & 13.46\% & ± 1.95\% \\
III ➜ IV & 15.74\% & ± 2.24\% \\
late T-cat. binom. prob. & 45.35\% & ± 2.57\% \\

@@

We want to investigate whether and to what extent the model can fulfill
the requirements laid out in section~\ref{sec-requirements}. To that
end, we compare its predictions for contralateral involvement against
observations in the data. This is done given scenarios that differ in
T-category and/or midline extension and/or ipsilateral involvement.

\subsubsection{Dependence of Contralateral Involvement on T-Category and
Midline
Extension}\label{dependence-of-contralateral-involvement-on-t-category-and-midline-extension}

@@

\subsubsection{Prevalence of Midline
Extension}\label{prevalence-of-midline-extension}

@@

@@

\caption{}\label{T_07ae0}\tabularnewline
\toprule\noalign{}
T-cat. & ipsi & Mid. ext. & \multicolumn{2}{l}{%
I} & \multicolumn{2}{l}{%
II} & \multicolumn{2}{l}{%
III} & \multicolumn{2}{l}{%
IV} & total \\
& & & n & \% & n & \% & n & \% & n & \% & n \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
T-cat. & ipsi & Mid. ext. & \multicolumn{2}{l}{%
I} & \multicolumn{2}{l}{%
II} & \multicolumn{2}{l}{%
III} & \multicolumn{2}{l}{%
IV} & total \\
& & & n & \% & n & \% & n & \% & n & \% & n \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
early & 0 & False & 0 & 0.00 & 1 & 1.16 & 0 & 0.00 & 0 & 0.00 & 86 \\
early & 0 & True & 0 & 0.00 & 1 & 10.00 & 1 & 10.00 & 0 & 0.00 & 10 \\
early & 1 & False & 1 & 0.53 & 11 & 5.82 & 2 & 1.06 & 1 & 0.53 & 189 \\
early & 1 & True & 1 & 11.11 & 2 & 22.22 & 0 & 0.00 & 0 & 0.00 & 9 \\
early & ≥ 2 & False & 1 & 0.96 & 15 & 14.42 & 3 & 2.88 & 3 & 2.88 &
104 \\
early & ≥ 2 & True & 0 & 0.00 & 3 & 30.00 & 4 & 40.00 & 1 & 10.00 &
10 \\
advanced & 0 & False & 0 & 0.00 & 2 & 5.88 & 0 & 0.00 & 0 & 0.00 & 34 \\
advanced & 0 & True & 0 & 0.00 & 3 & 12.50 & 0 & 0.00 & 0 & 0.00 & 24 \\
advanced & 1 & False & 0 & 0.00 & 3 & 4.55 & 0 & 0.00 & 0 & 0.00 & 66 \\
advanced & 1 & True & 1 & 1.64 & 18 & 29.51 & 5 & 8.20 & 1 & 1.64 &
61 \\
advanced & ≥ 2 & False & 4 & 5.26 & 16 & 21.05 & 7 & 9.21 & 4 & 5.26 &
76 \\
advanced & ≥ 2 & True & 3 & 3.80 & 46 & 58.23 & 18 & 22.78 & 8 & 10.13 &
79 \\

@@

  \bibliography{references.bib}
